---
title: "ERP"
author: "yyx"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: '3'
---

# Introduction

This report analyzes the P200, N400, and P600 components of event-related potentials (ERP) to investigate their amplitude and latency across different brain regions and word-class conditions. The workflow includes data preprocessing, repeated-measures ANOVA, and heatmap-based visualization.

# Load Required Packages

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ez)
```

# Read Data

Please update the file paths below to match your local environment.

```{r read-data}
data_p200_raw <- read.csv("/Users/yuanxiaoyu/Desktop/untitled folder/data/data_P200.csv",
                          stringsAsFactors = FALSE, check.names = FALSE)
data_n400_raw <- read.csv("/Users/yuanxiaoyu/Desktop/untitled folder/data/data_N400.csv",
                          stringsAsFactors = FALSE, check.names = FALSE)
data_p600_raw <- read.csv("/Users/yuanxiaoyu/Desktop/untitled folder/data/data_P600.csv",
                          stringsAsFactors = FALSE, check.names = FALSE)
```

# Standardize Text Columns

```{r clean-columns}
clean_basic_columns <- function(df) {
  df$position <- toupper(trimws(df$position))
  df$sample_name <- trimws(df$sample_name)
  if ("component" %in% names(df)) {
    df$component <- trimws(df$component)
  }
  return(df)
}

data_p200_raw <- clean_basic_columns(data_p200_raw)
data_n400_raw <- clean_basic_columns(data_n400_raw)
data_p600_raw <- clean_basic_columns(data_p600_raw)
```

# Convert Wide Format to Long Format

```{r wide-to-long}
convert_erp_wide_to_long <- function(df) {
  required_base_cols <- c("position", "sample_name", "component")
  missing_base_cols <- setdiff(required_base_cols, names(df))
  
  if (length(missing_base_cols) > 0) {
    stop(paste("Missing required base columns:", paste(missing_base_cols, collapse = ", ")))
  }
  
  measure_cols <- names(df)[grepl("_(μV|ms)$", names(df))]
  
  if (length(measure_cols) == 0) {
    stop("No measurement columns found ending in _μV or _ms")
  }
  
  long_df <- df %>%
    pivot_longer(
      cols = all_of(measure_cols),
      names_to = c("word_class", ".value"),
      names_pattern = "^(.*)_(μV|ms)$"
    ) %>%
    rename(
      amplitude = `μV`,
      latency = ms
    )
  
  long_df$word_class <- trimws(long_df$word_class)
  long_df$amplitude <- as.numeric(long_df$amplitude)
  long_df$latency <- as.numeric(long_df$latency)
  
  return(long_df)
}

data_p200 <- convert_erp_wide_to_long(data_p200_raw)
data_n400 <- convert_erp_wide_to_long(data_n400_raw)
data_p600 <- convert_erp_wide_to_long(data_p600_raw)
```

# Inspect Transformed Data

```{r inspect-data}
cat("P200 long-format columns:\n")
print(colnames(data_p200))
cat("\nP200 word classes:\n")
print(unique(data_p200$word_class))

cat("\nN400 long-format columns:\n")
print(colnames(data_n400))
cat("\nN400 word classes:\n")
print(unique(data_n400$word_class))

cat("\nP600 long-format columns:\n")
print(colnames(data_p600))
cat("\nP600 word classes:\n")
print(unique(data_p600$word_class))
```

# Define Functional Brain Regions for Each ERP Component

```{r define-brain-regions}
define_brain_regions_by_erp <- function() {
  brain_regions <- list(
    "P200" = list(
      "Frontal" = c("F3", "F4", "FZ", "F5", "F6"),
      "Frontocentral" = c("FC3", "FC4", "FCZ", "FC5", "FC6"),
      "Frontotemporal" = c("FT7", "FT8"),
      "Central" = c("C3", "C4", "CZ"),
      "Centroparietal" = c("CP3", "CP4", "CPZ"),
      "Occipital" = c("O1", "O2")
    ),
    "N400" = list(
      "Frontal" = c("F3", "F4", "FZ", "F5", "F7"),
      "Frontocentral" = c("FC3", "FC4", "FCZ", "FC5"),
      "Central" = c("C3", "C4", "CZ"),
      "Centroparietal" = c("CP3", "CP4", "CPZ"),
      "Parietal" = c("P3", "P4", "PZ"),
      "Occipital" = c("O1", "O2")
    ),
    "P600" = list(
      "Frontal" = c("F3", "F4", "FZ"),
      "Frontopolar" = c("FP1", "FP2"),
      "Frontocentral" = c("FC3", "FC4", "FCZ"),
      "Central" = c("C3", "C4", "CZ"),
      "Centroparietal" = c("CP3", "CP4", "CPZ"),
      "Parietal" = c("P3", "P4", "PZ", "P5", "P6"),
      "Temporal" = c("T7", "T8")
    )
  )
  return(brain_regions)
}

brain_regions <- define_brain_regions_by_erp()
```

# Define Electrode Positions

```{r define-positions}
p200_positions <- unique(unlist(brain_regions$P200))
n400_positions <- unique(unlist(brain_regions$N400))
p600_positions <- unique(unlist(brain_regions$P600))
```

# Map Electrodes to Brain Regions

```{r map-electrodes}
map_electrodes_to_regions <- function(data, erp_name) {
  brain_regions <- define_brain_regions_by_erp()
  erp_regions <- brain_regions[[erp_name]]
  
  position_to_region <- data.frame(
    position = character(),
    brain_region = character(),
    stringsAsFactors = FALSE
  )
  
  for (region_name in names(erp_regions)) {
    positions <- erp_regions[[region_name]]
    temp_df <- data.frame(
      position = positions,
      brain_region = region_name,
      stringsAsFactors = FALSE
    )
    position_to_region <- rbind(position_to_region, temp_df)
  }
  
  data_with_regions <- data %>%
    left_join(position_to_region, by = "position") %>%
    filter(!is.na(brain_region))
  
  return(data_with_regions)
}
```

# Aggregate by Brain Region

```{r aggregate-by-region}
aggregate_by_brain_region <- function(data, measure_col, erp_name) {
  data_with_regions <- map_electrodes_to_regions(data, erp_name)
  
  if (measure_col == "amplitude") {
    aggregated_data <- data_with_regions %>%
      group_by(sample_name, word_class, brain_region) %>%
      summarise(
        avg_amplitude = mean(amplitude, na.rm = TRUE),
        n_electrodes = n(),
        .groups = "drop"
      )
  } else {
    aggregated_data <- data_with_regions %>%
      group_by(sample_name, word_class, brain_region) %>%
      summarise(
        avg_latency = mean(latency, na.rm = TRUE),
        n_electrodes = n(),
        .groups = "drop"
      )
  }
  
  return(aggregated_data)
}
```

# Simplified Analysis

This fallback analysis is used when the full repeated-measures ANOVA cannot be completed successfully.

```{r simplified-analysis-functions}
run_simplified_analysis <- function(data, measure, erp_name) {
  cat("\nRunning Simplified Analysis - Main Effects Testing:\n")
  
  if (measure == "amplitude") {
    word_class_data <- data %>%
      group_by(sample_name, word_class) %>%
      summarise(mean_val = mean(avg_amplitude, na.rm = TRUE), .groups = "drop")
    
    brain_region_data <- data %>%
      group_by(sample_name, brain_region) %>%
      summarise(mean_val = mean(avg_amplitude, na.rm = TRUE), .groups = "drop")
  } else {
    word_class_data <- data %>%
      group_by(sample_name, word_class) %>%
      summarise(mean_val = mean(avg_latency, na.rm = TRUE), .groups = "drop")
    
    brain_region_data <- data %>%
      group_by(sample_name, brain_region) %>%
      summarise(mean_val = mean(avg_latency, na.rm = TRUE), .groups = "drop")
  }
  
  cat("\n--- Word Class Main Effect ---\n")
  tryCatch({
    word_anova <- ezANOVA(
      data = word_class_data,
      dv = mean_val,
      wid = sample_name,
      within = .(word_class),
      detailed = TRUE
    )
    print(word_anova$ANOVA)
  }, error = function(e) {
    cat("Word class effect analysis failed:", e$message, "\n")
  })
  
  cat("\n--- Brain Region Main Effect ---\n")
  tryCatch({
    region_anova <- ezANOVA(
      data = brain_region_data,
      dv = mean_val,
      wid = sample_name,
      within = .(brain_region),
      detailed = TRUE
    )
    print(region_anova$ANOVA)
  }, error = function(e) {
    cat("Brain region effect analysis failed:", e$message, "\n")
  })
  
  return(list(
    data = data,
    erp_name = erp_name,
    measure = measure,
    simplified = TRUE
  ))
}
```

# Main ERP Analysis Function

```{r main-analysis-function}
run_erp_analysis <- function(data, measure, erp_name, positions) {
  cat("\n", paste(rep("=", 60), collapse = ""), "\n", sep = "")
  cat("Analyzing", erp_name, "component", ifelse(measure == "amplitude", "Amplitude", "Latency"), "\n")
  cat(paste(rep("=", 60), collapse = ""), "\n", sep = "")
  
  data_filtered <- data %>%
    filter(position %in% positions)
  
  if (measure == "amplitude") {
    data_filtered <- data_filtered %>% filter(!is.na(amplitude))
  } else {
    data_filtered <- data_filtered %>% filter(!is.na(latency))
  }
  
  cat("Original data points:", nrow(data_filtered), "\n")
  cat("Number of subjects:", length(unique(data_filtered$sample_name)), "\n")
  cat("Number of electrode positions:", length(unique(data_filtered$position)), "\n")
  cat("Number of word classes:", length(unique(data_filtered$word_class)), "\n")
  
  aggregated_data <- aggregate_by_brain_region(data_filtered, measure, erp_name)
  
  cat("\nAggregated Data:\n")
  cat("Data points:", nrow(aggregated_data), "\n")
  cat("Brain regions:", paste(unique(aggregated_data$brain_region), collapse = ", "), "\n")
  cat("Word classes:", paste(unique(aggregated_data$word_class), collapse = ", "), "\n")
  
  aggregated_data$sample_name <- as.factor(aggregated_data$sample_name)
  aggregated_data$word_class <- as.factor(aggregated_data$word_class)
  aggregated_data$brain_region <- as.factor(aggregated_data$brain_region)
  
  n_regions <- length(unique(aggregated_data$brain_region))
  n_word_classes <- length(unique(aggregated_data$word_class))
  expected_per_subject <- n_regions * n_word_classes
  
  complete_check <- aggregated_data %>%
    group_by(sample_name) %>%
    summarise(
      n_obs = n(),
      expected = expected_per_subject,
      complete = (n_obs == expected),
      .groups = "drop"
    )
  
  complete_subjects <- complete_check$sample_name[complete_check$complete]
  cat("Subjects with complete data:", length(complete_subjects), "\n")
  
  if (length(complete_subjects) < 15) {
    cat("Insufficient complete subjects. Attempting to include subjects with at least 80% data...\n")
    partial_complete_subjects <- complete_check$sample_name[
      complete_check$n_obs >= 0.8 * expected_per_subject
    ]
    cat("Subjects with >= 80% data:", length(partial_complete_subjects), "\n")
    
    if (length(partial_complete_subjects) >= 15) {
      complete_subjects <- partial_complete_subjects
    }
  }
  
  final_data <- aggregated_data %>%
    filter(sample_name %in% complete_subjects) %>%
    droplevels()
  
  cat("Final number of subjects for analysis:", length(unique(final_data$sample_name)), "\n")
  cat("Final data points:", nrow(final_data), "\n")
  
  if (nrow(final_data) == 0) {
    cat("No valid data available for analysis after filtering.\n")
    return(list(
      data = NULL,
      anova = NULL,
      descriptive = NULL,
      erp_name = erp_name,
      measure = measure
    ))
  }
  
  if (measure == "amplitude") {
    descriptive_stats <- final_data %>%
      group_by(word_class, brain_region) %>%
      summarise(
        mean_amp = mean(avg_amplitude, na.rm = TRUE),
        sd_amp = sd(avg_amplitude, na.rm = TRUE),
        n = n(),
        .groups = "drop"
      )
    cat("\nDescriptive Statistics (Amplitude μV):\n")
    print(descriptive_stats, n = Inf)
  } else {
    descriptive_stats <- final_data %>%
      group_by(word_class, brain_region) %>%
      summarise(
        mean_lat = mean(avg_latency, na.rm = TRUE),
        sd_lat = sd(avg_latency, na.rm = TRUE),
        n = n(),
        .groups = "drop"
      )
    cat("\nDescriptive Statistics (Latency ms):\n")
    print(descriptive_stats, n = Inf)
  }
  
  tryCatch({
    cat("\nRunning Repeated Measures ANOVA...\n")
    
    if (measure == "amplitude") {
      anova_result <- ezANOVA(
        data = final_data,
        dv = avg_amplitude,
        wid = sample_name,
        within = .(word_class, brain_region),
        detailed = TRUE,
        type = 2
      )
    } else {
      anova_result <- ezANOVA(
        data = final_data,
        dv = avg_latency,
        wid = sample_name,
        within = .(word_class, brain_region),
        detailed = TRUE,
        type = 2
      )
    }
    
    cat("\nANOVA Results:\n")
    anova_table <- anova_result$ANOVA
    print(anova_table)
    
    cat("\nInterpretation of Significant Effects:\n")
    significant_effects <- anova_table[anova_table$p < 0.05, ]
    if (nrow(significant_effects) > 0) {
      for (i in 1:nrow(significant_effects)) {
        effect <- significant_effects$Effect[i]
        p_val <- significant_effects$p[i]
        f_val <- significant_effects$F[i]
        cat(sprintf("- %s: F = %.3f, p = %.3f (Significant)\n", effect, f_val, p_val))
      }
    } else {
      cat("- No significant effects found (p < 0.05)\n")
    }
    
    if (!is.null(anova_result$`Mauchly's Test for Sphericity`)) {
      cat("\nMauchly's Test for Sphericity:\n")
      mauchly_table <- anova_result$`Mauchly's Test for Sphericity`
      print(mauchly_table)
      
      sphericity_violations <- mauchly_table[mauchly_table$p < 0.05, ]
      if (nrow(sphericity_violations) > 0) {
        cat("\nSphericity Corrections (Greenhouse-Geisser):\n")
        print(anova_result$`Sphericity Corrections`)
      }
    }
    
    return(list(
      data = final_data,
      anova = anova_result,
      descriptive = descriptive_stats,
      erp_name = erp_name,
      measure = measure
    ))
    
  }, error = function(e) {
    cat("ANOVA failed:", e$message, "\n")
    cat("Attempting simplified analysis...\n")
    return(run_simplified_analysis(final_data, measure, erp_name))
  })
}
```

# Visualization Function

```{r visualization-function}
create_erp_plots <- function(analysis_result) {
  if (is.null(analysis_result$data)) return(NULL)
  
  data <- analysis_result$data
  erp_name <- analysis_result$erp_name
  measure <- analysis_result$measure
  
  if (measure == "amplitude") {
    plot_data <- data %>%
      group_by(word_class, brain_region) %>%
      summarise(mean_val = mean(avg_amplitude, na.rm = TRUE), .groups = "drop")
    
    p1 <- ggplot(plot_data, aes(x = word_class, y = brain_region, fill = mean_val)) +
      geom_tile() +
      scale_fill_gradient2(low = "blue", mid = "white", high = "red", name = "Amplitude (μV)") +
      theme_minimal() +
      labs(title = paste(erp_name, "Amplitude"), x = "Word Class", y = "Brain Region") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  } else {
    plot_data <- data %>%
      group_by(word_class, brain_region) %>%
      summarise(mean_val = mean(avg_latency, na.rm = TRUE), .groups = "drop")
    
    p1 <- ggplot(plot_data, aes(x = word_class, y = brain_region, fill = mean_val)) +
      geom_tile() +
      scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Latency (ms)") +
      theme_minimal() +
      labs(title = paste(erp_name, "Latency"), x = "Word Class", y = "Brain Region") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  }
  
  return(p1)
}
```

# Run the Full ERP Analysis

```{r wrapper-function}
run_complete_erp_analysis <- function() {
  results <- list()
  
  cat("Starting P200 Analysis...\n")
  results$p200_amplitude <- run_erp_analysis(data_p200, "amplitude", "P200", p200_positions)
  results$p200_latency   <- run_erp_analysis(data_p200, "latency", "P200", p200_positions)
  
  cat("\nStarting N400 Analysis...\n")
  results$n400_amplitude <- run_erp_analysis(data_n400, "amplitude", "N400", n400_positions)
  results$n400_latency   <- run_erp_analysis(data_n400, "latency", "N400", n400_positions)
  
  cat("\nStarting P600 Analysis...\n")
  results$p600_amplitude <- run_erp_analysis(data_p600, "amplitude", "P600", p600_positions)
  results$p600_latency   <- run_erp_analysis(data_p600, "latency", "P600", p600_positions)
  
  return(results)
}
```

# Execute All Analyses

```{r run-analysis}
cat("Starting ERP Analysis: Brain Regional Mapping vs Word Class Relationship\n")
cat(paste(rep("=", 80), collapse = ""), "\n", sep = "")

all_results <- run_complete_erp_analysis()
```

# Generate and Display Plots

```{r plot-results, fig.width=8, fig.height=5}
p_p200_amp <- create_erp_plots(all_results$p200_amplitude)
p_p200_lat <- create_erp_plots(all_results$p200_latency)

p_n400_amp <- create_erp_plots(all_results$n400_amplitude)
p_n400_lat <- create_erp_plots(all_results$n400_latency)

p_p600_amp <- create_erp_plots(all_results$p600_amplitude)
p_p600_lat <- create_erp_plots(all_results$p600_latency)

print(p_p200_amp)
print(p_p200_lat)
print(p_n400_amp)
print(p_n400_lat)
print(p_p600_amp)
print(p_p600_lat)
```

# Conclusion

This report provides a complete ERP analysis workflow for examining the relationship between brain-region functional mapping and word class across the P200, N400, and P600 components. The pipeline includes structured preprocessing, region-level aggregation, repeated-measures ANOVA, fallback simplified analysis, and heatmap visualization.
